Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-04-10 manuscript/manuscript.tex
more work on conclusion

Molecular phylogenetics is commonly used to estimate the historical timing of evolutionary events \citep{yang2012molecular}. 
However, phylogenetic estimates of the divergence times of many viral lineages are clearly too recent~\citep{duchene2014analyses, aiewsakun2016time}.
For example, the integration of filoviruses into their host genomes indicate that Ebola and Marburg virus diverged from their common ancestor 7 to 12 million years ago---but the estimate of this divergence time based on phylogenetic analyses of the viral sequences is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}. 
Similarly, the phylogenetic estimate of when major simian immunodeficiency virus groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of their host species~\citep{wertheim2009dating, worobey2010island}. 
These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin, rhazes1848treatise}\jdbcomment{add paper specifically emphasizing discrepancy}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.

@@

More recent work has used mutation-selection models to better account for purifying selection~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
These models explicitly account for the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\jdbcomment{cite any papers that show this is true. Maybe the CAT model and the one on long-branch attraction?}.
However, these approaches still ultimately involving inferring the site-specific purifying selection from natural sequence data.

@@

However, it is also clear that the amino-acid preferences defining the ExpCM's do not extrapolate well over a very diverged tree. 
It is well established that a protein's site-specific amino-acid preferences are contextually dependent~\citep{gong2013stability, harms2014historical, tufts2014epistasis, ortlund2007crystal} and shift as the protein accumulates substitutions overtime~\citep{haddox2018mapping, starr2018pervasive, pollock2012amino, doud2015site, shah2015contingency,bazykin2015changing}. 
The site-specific amino-acid preferences we used to define the ExpCM's were only measured in one genetic background. 
These preferences are only one snapshot of the functional constraint and do not capture any of the changes over time. 
The model's ignorance of shifting preferences is reflected in our results: The ExpCM's have the largest effect on branch length extension near the HA from the experiment.

@@

Clearly, the site-independence and time-homogeneity assumptions of the ExpCM's and pbMutSel models inhibit their ability to accurately describe the effect of shifting preferences. 
Models which account for context-dependent of amino-acid preferences have been developed but are still in their infancy. 
Currently, there are two main challenges associated with context-dependence models. 
The first challenge is how to efficiently evaluate these inherently more complex models mathematically so the models can be applied to large phylogenetic trees. 
The second challenge is developing the correct function to model shifting preferences. 
One strategy to to account for context-dependence is to evaluate amino-acid changes as how they might affect protein structure.
However, in nature a protein may undergo conformational changes in order to perform some function. 
Models which account for shifting preferences should also be flexible enough to be able to describe multiple, necessary protein conformations. 
Such models will be critical for describing the long-term evolution of a protein and accurately estimating the length of deep branches on phylogenetic trees.

@@

\subsection*{HA amino-acid preferences from deep mutational scanning experiments}
We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} to define the amino-acid preferences that inform the ExpCM's. 
We only used sites that can be unambiguously aligned in these H1 and H3 HAs. 
These alignable sites and their mapping to sequential numbering of the HA sequences used in the deep mutational scanning experiments are in \ref{suppfile:WSN_Perth_map}. 
The experimentally measured amino-acid preferences masked to just include these alignable sites are in \ref{suppfile:H1_prefs} and \ref{suppfile:H3_prefs}.
For the average preference set, we took the pairwise average of the H1 and H3 preferences. 
The preference for every amino acid $a$ at every site $r$ in the average preference set is
\begin{equation}
\label{eq:pi_avg}
\pi_{r,a,\rm{(H1+H3\ avg)}} = \frac{\pi_{r,a,\rm{H1}} + \pi_{r,a,\rm{H3}}}{2}
\end{equation}

@@

In order to make the branch lengths in \ref{fig:phylobayes} comparable between the pbMutSel tree returned by \texttt{PhyloBayes-MPI} and the other trees returned by \texttt{phydms}, we we normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006. 
These two H1 sequences are early and late representatives of the longest known human influenza lineage, and are of sufficiently high identity that different ExpCM and GY94 substitution models all estimate nearly identical branch lengths separating them.

>>>>

Molecular phylogenetics is commonly used to estimate the historical timing of evolutionary events \citep{yang2012molecular}. 
However, phylogenetic estimates of the divergence times of many viral lineages are clearly too recent~\citep{duchene2014analyses, aiewsakun2016time}.
For example, the integration of filoviruses into their host genomes indicate that Ebola and Marburg virus diverged from their common ancestor 7 to 12 million years ago---but the estimate of this divergence time based on phylogenetic analyses of the viral sequences is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}. 
Similarly, the phylogenetic estimate of when major simian immunodeficiency virus groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of their host species~\citep{wertheim2009dating, worobey2010island}. 
These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised, holmes2003molecular}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.

@@

More recent work has used mutation-selection models to better account for purifying selection~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
These models explicitly account for the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\citep{philippe1998good, lartillot2007suppression}.
However, these approaches still ultimately involving inferring the site-specific purifying selection from natural sequence data.

@@

However, it is also clear that the amino-acid preferences defining the ExpCM's do not extrapolate well over a very diverged tree. 
It is well established that a protein's site-specific amino-acid preferences are contextually dependent~\citep{gong2013stability, harms2014historical, tufts2014epistasis, ortlund2007crystal} and shift as the protein accumulates substitutions overtime~\citep{haddox2018mapping, starr2018pervasive, pollock2012amino, doud2015site, shah2015contingency,bazykin2015changing}. 
The site-specific amino-acid preferences we used to define the ExpCM's were only measured in one genetic background. 
Therefore, these preferences are only one snapshot of the functional constraint and do not capture any of the changes over time. 
The model's ignorance of shifting preferences is reflected in our results: The ExpCM's have the largest effect on branch length extension near the HA from the experiment.

@@

Clearly, the site-independence and time-homogeneity assumptions of the ExpCM's and pbMutSel models inhibit their ability to accurately describe the effect of shifting preferences. 
Models which account for context-dependent of amino-acid preferences have been developed but are still in their infancy.
Thus far their application has been restricted to a few sequences~\citep{rodrigue2005site, choi2007quantifying}, short evolutionary distances~\citep{bordner2013new}, or simulated sequences~\citep{goldstein2017sequence}.

Currently, there are two main challenges associated with context-dependence models. 
The first challenge is how to efficiently evaluate these inherently more complex models mathematically so the models can be applied to large phylogenetic trees~\citep{rodrigue2009computational}. 
The second challenge is developing the correct function to model shifting preferences. 
For example, \citet{rodrigue2009computational} showed that while site-dependent amino-acid models outperformed site-independent amino-acid models, they were outperformed by sophisticated site-independent codon models.
Most context-dependent models evaluate amino-acid changes as how they might affect a single protein conformation~\citep{rodrigue2005site, choi2007quantifying,bordner2013new, goldstein2017sequence}.
However, in nature a protein may undergo conformational changes in order to perform some function. 
Models which account for shifting preferences should also be flexible enough to be able to describe multiple, necessary protein conformations. 
Such models will be critical for describing the long-term evolution of a protein and accurately estimating the length of deep branches on phylogenetic trees.

@@

\subsection*{HA amino-acid preferences from deep mutational scanning experiments}
We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} \skhcomment{change citation} to define the amino-acid preferences that inform the ExpCM's. 
We only used sites that can be unambiguously aligned in these H1 and H3 HAs. 
These alignable sites and their mapping to sequential numbering of the HA sequences used in the deep mutational scanning experiments are in \ref{suppfile:WSN_Perth_map}. 
The experimentally measured amino-acid preferences masked to just include these alignable sites are in \ref{suppfile:H1_prefs} and \ref{suppfile:H3_prefs}.
For the average preference set, we took the pairwise average of the H1 and H3 preferences. 
The preference for every amino acid $a$ at every site $r$ in the average preference set is
\begin{equation}
\label{eq:pi_avg}
\pi_{r,a,\rm{(H1+H3\ avg)}} = \frac{\pi_{r,a,\rm{H1}} + \pi_{r,a,\rm{H3}}}{2}
\end{equation}

@@

In order to make the branch lengths in \ref{fig:phylobayes} comparable between the pbMutSel tree returned by \texttt{PhyloBayes-MPI} and the other trees returned by \texttt{phydms}, we we normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006. 
These two H1 sequences are early and late representatives of the longest known human influenza lineage, and are of sufficiently high identity that different ExpCM and GY94 substitution models all estimate nearly identical branch lengths separating them. 
\skhcomment{add new run parameters}
